{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-04-20 paper.tex
editing the introduction

\begin{abstract}
We present a new method conceived to ease the inclusion of hadronic
observables into parton distribution function (PDF) fits. Our
technique relies on the the same principles adopted by the existing
fast interfaces but implements a set of improvements aimed at
optimizing the performance in the context of PDF fits where many
iterations and many predictions for each iteration are usually
required. The main novelties are the precomputation of the PDF and
$\alpha_s$ evolution and the optimization of the numerical convolution
between hard cross sections and PDFs. We demostrate that our techique
can lead to a substantial speed improvement as compared to the
existing interfaces.
\end{abstract}

@@

The need for fast and accurate cross-section predictions in the Large
Hadron Collider (LHC) era is becoming increasingly important. In fact,
the outstanding amount of precise data produced by the LHC experiments
is bound to provide a very stringent constraint on the proton parton
distribution functions (PDFs) on the condition that as many
measurements as possible are included in PDF fits with the highest
accuracy possible. However, the direct use of most of the codes that
provide predictions for non-inclusive hadron-collider observables in a
PDF fit, which typically requires a large number of iterations, is not
viable because the time required to obtain accurate predictions is
usually of the order of a few hours or more. In order to overcome this
limitation, the typical strategy adopted relies on the precomputation
of the partonic hard cross sections in such a way that the numerical
convolution with any set of PDFs can be quickly performed {\it a
  posteriori} by means of interpolation techniques. Examples of the
implementation of this strategy are {\tt APPLgrid}~\cite{Carli:2010rw}
and {\tt FastNLO}~\cite{Wobisch:2011ij}. For the computation of the
hard cross sections, these packages rely on external codes to which
they are interfaced, more or less directly, by means of a suite of
functions whose scope is that of filling up PDF- and
$\alpha_s$-independent lookup tables of cross-section weights that can
subsequently be combined with an arbitrary PDF set. Examples of codes that have been directly interfaced to

\section{The FastKernel method}
Hadron collider observables are typically computed in QCD by a double convolution of parton densities with a hard scattering cross-section. Consider for example the calculation of a general cross section $pp\to X$ with a set of parton distribution functions $\{f\}$ up to order $N_\alpha$ in the strong coupling
\be  \sigma_{pp\to X} = \sum_{p}^{N_\alpha}\left( \frac{\alpha_s(Q^2)}{2\pi} \right)^{p+p_{\mathrm{LO}}} \int dx_1\, dx_2\,  f_i(x_1,Q^2)\; d\hat{\sigma}_{ij\to X}\; f_j(x_2,Q^2) \, \label{eq:hadconv} \ee
where $p_{\mathrm{LO}}$ is the leading order power of the strong coupling for the process $pp\to X$, $Q^2$ is the hard scale of the process and the indices $i,j$ sum over the active partonic species in the calculation. The central observation of tools such as {\tt APPLgrid} and {\tt fastNLO} is that such a calculation can be expressed as a more computationally efficient product

>>>>

\begin{abstract}
We present a new method conceived to ease the inclusion of hadronic
observables into parton distribution function (PDF) fits. Our
technique relies on the the same principles adopted by the existing
fast interfaces but implements a set of improvements aimed at
optimizing the performance in the context of PDF fits where many
iterations and many predictions for each iteration are usually
required. The main novelties are the precomputation of the PDF and
$\alpha_s$ evolution and the optimisation of the numerical convolution
between hard cross sections and PDFs. We demonstrate that our technique
can lead to a substantial speed improvement as compared to the
existing interfaces.
\end{abstract}

@@

The need for fast and accurate cross-section predictions in the Large
Hadron Collider (LHC) era is becoming increasingly important. In fact,
the outstanding amount of precise data produced by the LHC experiments
is bound to provide a very stringent constraint on the proton parton
distribution functions (PDFs) on the condition that as many
measurements as possible are included in PDF fits with the highest
accuracy possible. However, the direct use of most of the codes that
provide predictions for non-inclusive hadron-collider observables in a
PDF fit, which typically requires a large number of iterations, is not
viable because the time required to obtain accurate predictions is
usually of the order of a few hours or more. In order to overcome this
limitation, the typical strategy adopted relies on the precomputation
of the partonic hard cross sections in such a way that the numerical
convolution with any set of PDFs can be quickly performed {\it a
  posteriori} by means of interpolation techniques.

Examples of the implementation of this strategy are {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij}. For the computation of the hard
cross sections, these packages rely on external codes to which they
are interfaced, more or less directly, by means of a suite of
functions whose scope is that of filling up PDF- and
$\alpha_s$-independent look-up tables of cross-section weights that can
subsequently be combined with an arbitrary set of PDFs. Examples of
codes that have been directly interfaced to {\tt APPLgrid} and/or to
{\tt FastNLO} are {\tt MCFM}~\cite{Campbell:2010ff} and {\tt
  NLOJet++}~\cite{Nagy:2003tz}. More recently, dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. In particular, the {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators, respectively, and
fill up look-up interpolation tables in the {\tt APPLgrid} and/or {\tt
  FastNLO} format.

While these tools have proven to be extremely useful for the
extraction of parton densities, the great amount of experimental data
that has become available thanks to the LHC and that are or are going
to be included into the current PDF analyses is stretching the
capabilities of the general strategy as well as its
implementations. In fact, presently a typical global PDF fit includes
thousands of hadronic data points whose predictions have to be
computed thousands of times during the minimisation process. As a
consequence, the ``standard'' tools, $i.e.$ {\tt APPLgrid} and {\tt
  FastNLO}, might not be fast enough to meet the needs of the modern
PDF fitters. For this reason we believe that a high-performance tool
tailored for PDF analyses is becoming increasingly desirable.

The so-called {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was conceived to address
this problem. The basic principle of this method is the maximisation
of the precomputation with the consequent minimisation of the amount
of operations needed to be performed during the PDF fit. In
particular, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross sections and the DGLAP evolution kernels
and their combination into a single look-up table, which we call {\tt
  FKtable}, in such a way that the prediction for a given hadronic
observable can be obtained by a single numerical convolution between
the respective {\tt FKtable} and the initial scale PDFs, which are
the quantities that are usually fitted to data.

In this paper we present a public implementation of the {\tt
  FastKernel} method in which the partonic hard cross sections are
taken from any given {\tt APPLgrid} look-up table while the DGLAP
evolution kernels are provided by {\tt APFEL}
code~\cite{Bertone:2013vaa}. We named the resulting code {\tt
  APFELgrid}.

The outline of this paper is the following. In
Sect.~\ref{sec:FastKernel} we present the tecnical details of the
implemantation of the {\tt FastKernel} method.

\section{The FastKernel method}\label{sec:FastKernel}

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering
cross-section. Consider for example the calculation of a general cross
section $pp\to X$ with a set of parton distribution functions $\{f\}$
up to order $N_\alpha$ in the strong coupling \be \sigma_{pp\to X} =
\sum_{p}^{N_\alpha}\left( \frac{\alpha_s(Q^2)}{2\pi}
\right)^{p+p_{\mathrm{LO}}} \int dx_1\, dx_2\, f_i(x_1,Q^2)\;
d\hat{\sigma}_{ij\to X}\; f_j(x_2,Q^2) \, \label{eq:hadconv} \ee where
$p_{\mathrm{LO}}$ is the leading order power of the strong coupling
for the process $pp\to X$, $Q^2$ is the hard scale of the process and
the indices $i,j$ sum over the active partonic species in the
calculation. The central observation of tools such as {\tt APPLgrid}
and {\tt FastNLO} is that such a calculation can be expressed as a
more computationally efficient product \be
